\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-04 paper.tex
Move implementation and efficiency back into main text
Partial recordering of appendixes

The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% This is a digression from the maing point of the paragraph, so
% making it a parenthetical aside.
It is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The figure shows the pedigree of eight diploid individuals
and their sixteen constituent genomes (each consisting of a single chromosome),
along with paths of genetic inheritance.
Here, and throughout,
nodes are labelled with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic intervals.
Thus individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the two genomes belonging to an individual parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees along the
genome (C).
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes;
see Section~\ref{sec-ARG-and-local-trees} for details
on how local trees are embedded in an ARG.

@@

\section{Event ARGs}
\label{sec-eARG}
A common, classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This eARG encoding is the basis of the output formats created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this case, two types of internal node exist in the graph. These
represent the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

There is also a certain clarity gained by explicitly modelling nodes
in the inheritance graph as genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occurred,
or whether multiple coalescences
within a single individual should be regarded as one or multiple events are
debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
The gARG format avoids the need for such considerations,
so is more naturally compatible with forming an ecosystem of interoperable
inference and analysis methods.

@@

@@

\section{Survey of ARG inference methods}
\label{sec-survey-arg-infer}

The problem of reconstructing ARGs for samples of recombining sequences has
been of interest since the ARG was first defined. Early methods focused on
finding \emph{parsimonious} ARGs, i.e.\ those with a minimal number of
recombination events~\citep{hein1990reconstructing}. Two main approaches
emerged: ``backwards-in-time''~\citep{lyngso2005minimum} and
``along-the-genome'' \citep{song2003parsimonious, song2005constructing}.
Backwards-in-time approaches
start with a data matrix and reduce it to an empty matrix through row
and column operations corresponding to coalescence, mutation, and recombination
events, which construct an ARG from the bottom up~\citep{song2005efficient,
wu2008association,thao2019hybrid,ignatieva2021kwarg}.
Along-the-genome approaches
begin from an initial local tree at a single focal site.
Moving the focal site along the
genome changes the local tree via a subtree prune and regraft
operation whenever a recombination is encountered~\citep{hein1993heuristic,
wu2011new, mirzaei2017rent}.
\citet{rasmussen2022espalier} focus on parsimonious fusion of local
trees into an ARG, while the method described by
\citet{camara2016inference} is based on topological data analysis.
% The "most" is understood here, surely
Reconstructing a parsimonious ARG for a given data set is NP-hard
\citep{wang2001perfect}, so parsimony-based methods resort to heuristics and
are limited to analysing at most hundreds of sequences. Hence, a number of
methods aim to balance computational efficiency with reconstruction of
``reasonable'', rather than parsimonious ARGs
% distribution under a population genetics model
\citep{minichiello2006mapping,
parida2008estimating, kelleher2019inferring,  speidel2019method,
schaefer2021ancestral, zhang2023biobank}.

@@

%% JK:  What does this add that is not in caption?
% Thick black lines trace an ARG that describes the ancestry of the lower arm of
% the four sampled chromosomes, and these lines pass through gARG nodes (green
% chromosomal regions). A number of these nodes have only one parent and one
% child in the ARG; these would often be omitted (taken as implicit) in a
% minimally simplified gARG.

% Note that mutations can occur along any lineage in Fig.~\ref{fig-cell-lines}. This means, for example,
% that a mutation in the first cell division of $D_{10}$ could be shared between
% the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
% the right hand gamete. With a high enough mutation rate (or equivalently, a long enough genome), each
% round of mitotic germline genome duplication within a single multicellular organism could
% potentially be distinguished.
% This implies that in ARGs that have been inferred from real data, polytomies represent a lack
% of mutational information (rather than marking a single ``multifurcation event'').
% The ability to represent polytomies, and conceptualise them as a (deliberately) imprecise statement about the order
% of coalescence within and between individuals,
% fits naturally into the gARG framework presented in this paper.

>>>>

The breakthroughs in scale achieved by recent
methods are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history
(see Appendix~\ref{sec-survey-arg-infer} for a brief survey).
It is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The figure shows the pedigree of eight diploid individuals
and their sixteen constituent genomes (each consisting of a single chromosome),
along with paths of genetic inheritance.
Here, and throughout,
nodes are labelled with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic intervals.
Thus individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the two genomes belonging to an individual parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees along the
genome (C).
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes;
see Appendix~\ref{sec-ARG-and-local-trees} for details
on how local trees are embedded in an ARG.

@@

\section{Event ARGs}
\label{sec-eARG}
A common, classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This eARG encoding is the basis of the output formats created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding there are two types of internal node in the graph. These
represent the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

There is also a certain clarity gained by explicitly modelling nodes
in the inheritance graph as genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occurred,
or whether multiple coalescences
within a single individual should be regarded as one or multiple events are
debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
% The gARG formulation avoids the need for such considerations,
% so is more naturally compatible with forming an ecosystem of interoperable
% inference and analysis methods.

@@

\section{Survey of ARG inference methods}
\label{sec-survey-arg-infer}
The problem of reconstructing ARGs for samples of recombining sequences has
been of interest since the ARG was first defined. Early methods focused on
finding \emph{parsimonious} ARGs, i.e.\ those with a minimal number of
recombination events~\citep{hein1990reconstructing}. Two main approaches
emerged: ``backwards-in-time''~\citep{lyngso2005minimum} and
``along-the-genome'' \citep{song2003parsimonious, song2005constructing}.
Backwards-in-time approaches
start with a data matrix and reduce it to an empty matrix through row
and column operations corresponding to coalescence, mutation, and recombination
events, which construct an ARG from the bottom up~\citep{song2005efficient,
wu2008association,thao2019hybrid,ignatieva2021kwarg}.
Along-the-genome approaches
begin from an initial local tree at a single focal site.
Moving the focal site along the
genome changes the local tree via a subtree prune and regraft
operation whenever a recombination is encountered~\citep{hein1993heuristic,
wu2011new, mirzaei2017rent}.
\citet{rasmussen2022espalier} focus on parsimonious fusion of local
trees into an ARG, while the method described by
\citet{camara2016inference} is based on topological data analysis.
% The "most" is understood here, surely
Reconstructing a parsimonious ARG for a given data set is NP-hard
\citep{wang2001perfect}, so parsimony-based methods resort to heuristics and
are limited to analysing at most hundreds of sequences. Hence, a number of
methods aim to balance computational efficiency with reconstruction of
``reasonable'', rather than parsimonious ARGs
% distribution under a population genetics model
\citep{minichiello2006mapping,
parida2008estimating, kelleher2019inferring,  speidel2019method,
schaefer2021ancestral, zhang2023biobank}.

@@

@@
